Open Data, Open Source and Open Standards in chemistry - The Blue Obelisk five years on

### 2011-04-14 paper.tex
Fix spellings

Open Source toolkits for cheminformatics have now existed for nearly
ten years. During this period, some toolkits were developed form
scratch, whereas others were made Open Source by releasing in-house
codebases under liberal licenses. The primary toolkits currently in
active development are the Chemistry Development Kit, OpenBabel and
RDKit. These projects support a variety of languages natively (Java,
C++, Python), but by virtue of language wrappers, they can easily be
used from languages that they were not originally developed in. For
example, the CDK can be accessed from Python and OpenBabel can be
accesed from Java via SWIG wrappers. One important aspect of the
toolkits noted here is that they originated for the support of desktop
and server-side applications.

@@

In terms of functionality, most Open Source libraries support commonly
ued cheminformatics features. At the same time, individual toolkits
have their own strenghts. For example, the CDK has an extensive set of
molecular descriptors useful for QSAR modeling applictions, but lacks
support for 3D structure generation. In contrast OpenBabel, supports
3D structure generation and optimization using a variety of
forcefields, but lacks a built in set of molecular descriptors. RDKit
on the other hand supports both features and in general is the most
comprehensive Open Source toolkit in terms of functionality. Table XXX
summarizes the features of these three toolkits.

@@

An example of such an interaction was the donation by eMolecules, Inc. to Open Babel of code for the canonicalisation of molecules and fragments (Nov 2006). eMolecules is an online vendor of chemicals that uses Open Babel under-the-hood to manage its compound collections. Another more recent example occurred in July 2010 when Silicos, a Belgian company that provides services
in the area of cheminformatics, released several command line applications based on Open Babel as well as donating code to the project. For example, the Pharao
tool released by them is a comprehensive solution for pharmacophore
searching that provides extensive support for a variety of
pharmacophore searches. Other tools released by them include Seive for
fitering by molecular property and XXX.

@@

A core requirement for chemical structure databases and chemical
registration systems in general is the notion of structure
standardization.  That is,  for a given input structure, multiple
representations should be converted to one canonical form. 
Structure canonicalization routines partially address this aspect,
converting multiple alternative topologies to a single canonical
form. However, the problem of standardization is broader than just
topological canonicalization. Features that must be considered include
\begin{itemize}
\item topological canonicalization
\item handling of charges
\item tautomer enumeration and canonicalization
\item normalization of functional groups
\end{itemize}
Currently, most of the individual components of a `standardization
pipeline' can be implemented using BO tools. The larger problem is
that there is no agreed upon list of steps for a standardization
process. While some specifications have been published (e.g., Pubchem)
and some standardization services and tools are available (Pubchem
provides an online service to standaridize molecules and the NCGC
provides a stand alone tool) each group has their own set of rules. A
common reference specification for standardization would be of immense
value in interoperability between structure repositories as well as
between toolkits (though the latter is still confounded by differences
in lower level cheminformatic features such as aromaticity models).

>>>>

Open Source toolkits for cheminformatics have now existed for nearly
ten years. During this period, some toolkits were developed form
scratch, whereas others were made Open Source by releasing in-house
codebases under liberal licenses. The primary toolkits currently in
active development are the Chemistry Development Kit, Open Babel and
RDKit. These projects support a variety of languages natively (Java,
C++, Python), but by virtue of language wrappers, they can easily be
used from languages that they were not originally developed in. For
example, the CDK can be accessed from Python and Open Babel can be
accessed from Java via SWIG wrappers. One important aspect of the
toolkits noted here is that they originated for the support of desktop
and server-side applications.

@@

In terms of functionality, most Open Source libraries support commonly
used cheminformatics features. At the same time, individual toolkits
have their own strengths. For example, the CDK has an extensive set of
molecular descriptors useful for QSAR modeling applications, but lacks
support for 3D structure generation. In contrast, Open Babel supports
3D structure generation and optimization using a variety of
forcefields, but lacks a built in set of molecular descriptors. RDKit
on the other hand supports both features and in general is the most
comprehensive Open Source toolkit in terms of functionality. Table XXX
summarizes the features of these three toolkits.

@@

An example of such an interaction was the donation by eMolecules, Inc. to Open Babel of code for the canonicalisation of molecules and fragments (Nov 2006). eMolecules is an online vendor of chemicals that uses Open Babel under-the-hood to manage its compound collections. Another more recent example occurred in July 2010 when Silicos, a Belgian company that provides services
in the area of cheminformatics, released several command line applications based on Open Babel as well as donating code to the project. For example, the Pharao
tool released by them is a comprehensive solution for pharmacophore
searching that provides extensive support for a variety of
pharmacophore searches. Other tools released by them include Sieve for
filtering by molecular property and XXX.

@@

A core requirement for chemical structure databases and chemical
registration systems in general is the notion of structure
standardization.  That is,  for a given input structure, multiple
representations should be converted to one canonical form. 
Structure canonicalization routines partially address this aspect,
converting multiple alternative topologies to a single canonical
form. However, the problem of standardization is broader than just
topological canonicalization. Features that must be considered include
\begin{itemize}
\item topological canonicalization
\item handling of charges
\item tautomer enumeration and canonicalization
\item normalization of functional groups
\end{itemize}
Currently, most of the individual components of a `standardization
pipeline' can be implemented using BO tools. The larger problem is
that there is no agreed upon list of steps for a standardization
process. While some specifications have been published (e.g., Pubchem)
and some standardization services and tools are available (Pubchem
provides an online service to standardise molecules and the NCGC
provides a stand alone tool) each group has their own set of rules. A
common reference specification for standardization would be of immense
value in interoperability between structure repositories as well as
between toolkits (though the latter is still confounded by differences
in lower level cheminformatic features such as aromaticity models).
